---
title: "Cleaning administrative spatial data"
vignette: >
  %\VignetteIndexEntry{Cleaning administrative spatial data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    df-print: default
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    message: false
    warning: false
---

::: callout-note
Every layer in this vignette is **synthetic** — a toy world of rectangles with
fabricated names, built on the fly and deliberately broken so you can watch the
cleaner repair it. No real boundary data is used.
:::

## What `process_spatial()` is for

WHO ships administrative boundaries (country `ADM0`, province `ADM1`, district
`ADM2`) as spatial layers — an Esri geodatabase, a GeoPackage, or a folder of
shapefiles. Before that geography can drive analysis it has to be made
trustworthy, and raw boundary exports rarely are. The same export will typically
carry:

- **invalid geometries** — self-intersections and bad ring topology that break
  spatial joins;
- **sliver artifacts** — hairline holes and tiny detached fragments left by
  digitising, which are not real geography;
- **open-ended validity windows** — an `enddate` set to a sentinel year (e.g.
  `9999`) meaning "still current";
- **duplicate** units, and admin **names that vary** across releases.

`polished::process_spatial()` is a single call that reads each level, standardises
its columns to the package's canonical `adm0` / `adm1` / `adm2` names, audits it,
repairs the geometry, and writes both a cleaned per-level shape and a
year-expanded "long" shape for temporal joins.

## A folder of layers, deliberately broken

We build three layers — country, province, district — and write them to a
temporary folder as GeoPackages, exactly the shape `process_spatial()` expects to
find on disk. The district layer carries the interesting problems.

```{r}
#| label: build-synthetic
#| code-fold: true
# small geometry constructors -------------------------------------------------
rect <- function(xmin, ymin, xmax, ymax) {
  sf::st_polygon(list(rbind(
    c(xmin, ymin), c(xmin, ymax), c(xmax, ymax), c(xmax, ymin), c(xmin, ymin)
  )))
}
ring <- function(xmin, ymin, xmax, ymax) {
  rbind(c(xmin, ymin), c(xmin, ymax), c(xmax, ymax), c(xmax, ymin), c(xmin, ymin))
}

# ADM2 districts, each with a deliberate flaw -------------------------------
# D1: a real interior hole (large) PLUS a hairline sliver hole (tiny)
d1 <- sf::st_polygon(list(
  ring(0, 0, 1, 1), # outer
  ring(0.3, 0.3, 0.7, 0.7), # real hole  -> kept
  ring(0.9, 0.9, 0.9005, 0.9005) # sliver hole -> removed
))
# D2: a self-intersecting "bowtie" -> invalid, repaired by st_make_valid()
d2 <- sf::st_polygon(list(rbind(
  c(2, 0), c(3, 1), c(3, 0), c(2, 1), c(2, 0)
)))
# D3: a multipolygon = one real part + one tiny detached sliver part
d3 <- sf::st_multipolygon(list(
  list(ring(4, 0, 5, 1)), # real part   -> kept
  list(ring(6, 0, 6.0005, 6.0005)) # sliver part -> removed
))
# D4 and its exact duplicate (same name + geometry) -> flagged
d4 <- rect(0, 2, 1, 3)

shp_adm2 <- sf::st_sf(
  adm0_name = "ALPHA",
  adm0_guid = "g-alpha",
  adm1_name = c("NORTH", "NORTH", "SOUTH", "SOUTH", "SOUTH"),
  adm1_guid = c("g-an", "g-an", "g-as", "g-as", "g-as"),
  adm2_name = c("D1", "D2", "D3", "D4", "D4"),
  guid = c("g-d1", "g-d2", "g-d3", "g-d4", "g-d4"),
  startdate = "2015-01-01",
  enddate = c("9999-12-31", "9999-12-31", "2020-12-31", "9999-12-31", "9999-12-31"),
  geometry = sf::st_sfc(d1, d2, d3, d4, d4, crs = 4326)
)

# ADM1 provinces and ADM0 country (dissolved extents, no flaws) ---------------
shp_adm1 <- sf::st_sf(
  adm0_name = "ALPHA", adm0_guid = "g-alpha",
  adm1_name = c("NORTH", "SOUTH"), guid = c("g-an", "g-as"),
  startdate = "2015-01-01", enddate = "9999-12-31",
  geometry = sf::st_sfc(rect(0, 0, 1, 1), rect(2, 0, 6, 3), crs = 4326)
)
shp_adm0 <- sf::st_sf(
  adm0_name = "ALPHA", guid = "g-alpha",
  startdate = "2015-01-01", enddate = "9999-12-31",
  geometry = sf::st_sfc(rect(0, 0, 6, 3), crs = 4326)
)

# write the three layers into one folder --------------------------------------
src_dir <- file.path(tempdir(), "boundaries")
dir.create(src_dir, showWarnings = FALSE)
sf::st_write(shp_adm0, file.path(src_dir, "world_adm0.gpkg"), quiet = TRUE, delete_dsn = TRUE)
sf::st_write(shp_adm1, file.path(src_dir, "world_adm1.gpkg"), quiet = TRUE, delete_dsn = TRUE)
sf::st_write(shp_adm2, file.path(src_dir, "world_adm2.gpkg"), quiet = TRUE, delete_dsn = TRUE)
```

Here is the district layer the cleaner will be handed — five rows, four
distinct districts, and a catalogue of problems:

```{r}
#| label: messy-input
shp_adm2 |>
  sf::st_drop_geometry() |>
  dplyr::select(adm1_name, adm2_name, guid, enddate)
```

```{r}
#| label: messy-diagnostics
is_multipart <- function(geom) {
  vapply(geom, \(g) inherits(g, "MULTIPOLYGON") && length(g) > 1L, logical(1))
}

tibble::tibble(
  district = shp_adm2$adm2_name,
  valid = sf::st_is_valid(shp_adm2),
  multipart = is_multipart(sf::st_geometry(shp_adm2)),
  open_ended = lubridate::year(as.Date(shp_adm2$enddate)) == 9999
)
```

`D2` is invalid (the bowtie); `D1` carries a hairline sliver hole and `D3` a tiny
detached sliver part; `D4` is duplicated; and four of the five rows are
open-ended (`enddate` year `9999`).

## One call

Point `process_spatial()` at the folder, map each admin level to the filename
stem of its layer, and pick an output format. Repair is on by default.

```{r}
#| label: the-call
out_dir <- file.path(tempdir(), "boundaries_clean")

polished::process_spatial(
  input_path = src_dir,
  output_dir = out_dir,
  layers = c(
    adm0 = "world_adm0",
    adm1 = "world_adm1",
    adm2 = "world_adm2"
  ),
  output_format = "rds",
  verbose = FALSE
)
```

In an interactive session the default `verbose = TRUE` prints a per-level
summary of what was repaired (invalid geometries fixed, sliver holes and parts
removed); we silence it here. The lasting output is the set of files it writes:

```{r}
#| label: outputs
list.files(out_dir, recursive = TRUE)
```

## What it writes

<div class="desc-table">

| File | What it is |
|---|---|
| `spatial_global_{level}.rds` | the cleaned, repaired, reprojected layer |
| `spatial_{level}_long_shape.rds` | one row per unit **per active year** (ADM1/ADM2) |
| `checks/spatial_*.csv` | any validity / duplicate issues found, for review |

</div>

<details>
<summary><code>spatial_global_adm2.rds</code> — the cleaned layer</summary>

The cleaned district layer carries the canonical `adm0` / `adm1` / `adm2` names
(renamed from the source `*_name` columns), the level GUID as `adm2_guid`, and a
`shape` geometry column. Every geometry is now valid:

```{r}
#| label: cleaned-adm2
clean_adm2 <- readRDS(file.path(out_dir, "spatial_global_adm2.rds"))

clean_adm2 |>
  sf::st_drop_geometry() |>
  dplyr::select(adm0, adm1, adm2, adm2_guid)

# every geometry is valid after repair
all(sf::st_is_valid(clean_adm2))
```

</details>

<details>
<summary><code>spatial_adm2_long_shape.rds</code> — expanded by year</summary>

The long shape turns each unit's `year_start`..`year_end` validity window into
one row per active year — the table you join point data against when you need
the boundary that was current *in a given year*. Two rules keep it compact:

- enumeration stops at the **current year** (open-ended `9999` units do not
  spawn thousands of phantom rows); and
- a sentinel **`9999`** year is appended to every unit, so records whose year is
  unknown still match.

```{r}
#| label: long-shape
long_adm2 <- readRDS(file.path(out_dir, "spatial_adm2_long_shape.rds"))

# D3 closed in 2020; the open-ended districts run to the present + the 9999 key
long_adm2 |>
  dplyr::filter(adm2 %in% c("D3", "D1")) |>
  dplyr::group_by(adm2) |>
  dplyr::summarise(
    first_year = min(active_year),
    last_real_year = max(active_year[active_year != 9999]),
    has_sentinel = 9999 %in% active_year,
    .groups = "drop"
  )
```

</details>

<details>
<summary><code>checks/</code> — what was flagged for review</summary>

Problems are written to CSVs rather than silently dropped. The duplicated `D4`
district, for instance, lands in a duplicate-name check:

```{r}
#| label: checks
list.files(file.path(out_dir, "checks"))

readr::read_csv(
  file.path(out_dir, "checks", "spatial_duplicate_adm2_name.csv"),
  show_col_types = FALSE
) |>
  dplyr::select(adm0, adm1, adm2)
```

</details>

## Geometry repair

With `fix_issues = TRUE` (the default), `process_spatial()` does more than flag —
it repairs the output geometry through four steps:

<div class="desc-table">

| Step | What it does | Safe because… |
|---|---|---|
| drop Z/M | strips elevation/measure to clean 2D | admin work is planar |
| `st_make_valid()` | fixes self-intersections and ring topology | preserves the shape |
| remove sliver **holes** | drops interior rings below `sliver_area` | real enclaves/lakes are larger and kept |
| remove sliver **parts** | drops detached fragments below `sliver_area` | a feature never loses its **largest** part |

</div>

The two sliver steps are **threshold-based**: only artifacts smaller than
`sliver_area` (default `1e4` m² = 1 hectare) are removed. A genuine enclave, lake
or island above the threshold is left untouched, and no feature is ever emptied —
each keeps at least its largest polygon. You can see all three repairs land on
the district layer:

```{r}
#| label: repair-evidence
# D1 is the first feature; count its interior rings (holes)
d1_holes <- \(layer) length(sf::st_geometry(layer)[[1]]) - 1L

sum(!sf::st_is_valid(shp_adm2)) # one invalid geometry going in (the D2 bowtie)
sum(!sf::st_is_valid(clean_adm2)) # zero coming out
d1_holes(shp_adm2) # D1 had two holes (one real, one sliver)
d1_holes(clean_adm2) # ...and keeps the real one; only the sliver went
```

Set `fix_issues = FALSE` for the fastest, flag-only run (issues are still written
to `checks/`, but the geometry is left exactly as read), or tune `sliver_area`
up or down if one hectare is too aggressive or too lenient for your data.

## Reading from different sources

`input_path` is format-agnostic — whatever [`sf::st_read()`][sf::st_read] can
open. `layers` maps each admin level to a **layer name** (inside a `.gdb` /
`.gpkg`) or a **filename stem** (in a folder of standalone files), matched
case-insensitively.

<div class="desc-table">

| `input_path` is… | `layers` values match… |
|---|---|
| an Esri geodatabase (`.gdb`) | layer names inside the geodatabase |
| a GeoPackage (`.gpkg`) holding all levels | layer names inside the GeoPackage |
| a folder of shapefiles (`.shp`) | the `.shp` filename stems |
| a folder of GeoJSON (`.geojson`) | the `.geojson` filename stems |

</div>

`layers` may also be a **subset** of `adm0` / `adm1` / `adm2` — only the levels
you name are processed, so a single entry cleans one level:

```{r}
#| label: single-level
#| eval: false
# clean only the district layer
polished::process_spatial(
  input_path = src_dir,
  output_dir = out_dir,
  layers = c(adm2 = "world_adm2")
)
```

## Expanding a shape by year on its own

The year-expansion is also exported as `create_long_shape()`, useful when you
already have a cleaned layer and just want the per-year table. It collapses each
unit to a single validity window, then expands it:

```{r}
#| label: create-long-shape
provinces <- tibble::tibble(
  adm0_guid = "g-alpha", adm0 = "ALPHA",
  adm1_guid = c("g-an", "g-as"), adm1 = c("NORTH", "SOUTH"),
  year_start = c(2015, 2018), year_end = c(9999, 2021)
)

polished::create_long_shape(provinces, level = "adm1") |>
  dplyr::group_by(adm1) |>
  dplyr::summarise(
    years = paste0(min(active_year), "–", max(active_year[active_year != 9999])),
    sentinel = 9999 %in% active_year,
    .groups = "drop"
  )
```

`NORTH` is open-ended (`9999`), so it runs from 2015 to the current year;
`SOUTH` closed in 2021. Both gain the `9999` catch-all key.

## Recovering admin geography for point data

The third exported helper, `get_admin_info_from_coords()`, goes the other way:
given points with longitude/latitude, it spatially joins them to a district
layer and fills in any missing admin names and GUIDs. Points that fall on a
boundary and match more than one district are dropped rather than guessed.

```{r}
#| label: coords-join
points <- tibble::tibble(
  site = c("clinic-1", "clinic-2", "clinic-3"),
  adm0 = "ALPHA",
  adm1 = NA_character_,
  adm2 = NA_character_,
  adm1_guid = NA_character_,
  adm2_guid = NA_character_,
  longitude = c(0.1, 4.2, 0.8), # all three fall inside a clean district
  latitude = c(0.9, 0.5, 0.1) # D1, D3, D1
)

polished::get_admin_info_from_coords(points, clean_adm2) |>
  dplyr::select(site, adm0, adm1, adm2)
```

Each clinic inherits the province and district of the polygon it falls inside —
`adm1` and `adm2` were blank on the way in and are recovered from the cleaned
boundary layer.

## Parameters at a glance

<div class="desc-table">

| Argument | Default | Controls |
|---|---|---|
| `input_path` | — | the `.gdb` / `.gpkg` / folder to read |
| `output_dir` | — | where cleaned shapes and the `checks/` folder are written |
| `layers` | `GLOBAL_ADM*` | level → layer-name / filename-stem map (subset allowed) |
| `transform` | `TRUE` | reproject to `crs` (skipped when already equivalent) |
| `crs` | `4326` | target CRS (EPSG code) |
| `fix_issues` | `TRUE` | repair geometries (make-valid + slivers) vs flag-only |
| `sliver_area` | `1e4` | m² threshold below which holes / parts are artifacts |
| `output_format` | `"rds"` | serialise cleaned shapes as `"rds"` or `"qs2"` |
| `verbose` | `TRUE` | print the per-level cli summary |

</div>

For large global layers, `"qs2"` writes and reads the geometry far faster than
`"rds"` (it needs the `qs2` package), and `fix_issues = FALSE` is the quickest
flag-only pass when you already trust the geometry.
