---
title: "Cleaning POLIS data"
vignette: >
  %\VignetteIndexEntry{Cleaning POLIS data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    df-print: kable
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    message: false
    warning: false
---

```{r}
#| label: setup
#| include: false
library(polished)
```

The cleaning layer turns the raw POLIS tables into canonical, analysis-ready
tables — one cleaner per surveillance stream, all following the same recipe:
standardise names, sanitise dates, derive the analytic variables, clean the
geography, enrich, and deduplicate to one row per POLIS `id`.

| Function | Stream | Built from |
|---|---|---|
| `clean_afp()` | AFP cases | the Case table |
| `clean_human_spec()` | human lab specimens | the LabSpecimen table |
| `clean_es()` | environmental samples | the EnvSamples table |
| `clean_sia()` | immunisation campaigns | the Activity + SubActivity tables |
| `clean_virus()` | poliovirus positives | the **cleaned** AFP + ES outputs |

Two functions tie it together: `polis_config()` holds the settings every cleaner
shares, and `run_pipeline()` runs the whole set in one call.

The examples below use small synthetic frames with raw POLIS-style column names
(no spaces) so they render without a live POLIS connection; on real data you
pass the downloaded tables straight in.

## Shared settings: `polis_config()`

Every cleaner takes a `cfg` argument. `polis_config()` builds it — a single,
overridable object holding the column crosswalk, the output column ordering, and
run constants. There are deliberately no per-column arguments: the canonical
schema lives in the packaged crosswalk, not in scattered defaults.

```{r}
#| label: config
cfg <- polis_config(start_year = 2018)
cfg

# the column-ordering convention every cleaner emits to
names(cfg$column_roles)
```

The crosswalk that maps raw POLIS API names to canonical `snake_case` names is
shipped with the package and inspectable directly:

```{r}
#| label: crosswalk
head(polis_crosswalk()[, c("Table", "API_Name", "Snake_Name")])
```

## AFP cases: `clean_afp()`

`clean_afp()` standardises the Case table and derives the case-level analytic
variables AFP surveillance relies on: onset year/age, the onset-relative date
intervals, stool timeliness and 60-day follow-up flags, the fused virus
classification (`classification_all`), and the country enrichment.

```{r}
#| label: afp-raw
afp_raw <- data.frame(
  Id = c(1, 1, 2),
  Epid = c("NIE-BOR-MMC-24-001", "NIE-BOR-MMC-24-001", "NIE-BOR-JER-24-014"),
  LastUpdateDate = c("2024-02-01", "2024-04-01", "2024-03-01"),
  ParalysisOnsetDate = c("2024-01-02", "2024-01-02", "2024-02-03"),
  NotificationDate = c("2024-01-05", "2024-01-05", "2024-02-06"),
  Stool1CollectionDate = c("2024-01-08", "2024-01-08", "2024-02-10"),
  Stool2CollectionDate = c("2024-01-10", "2024-01-10", "2024-02-13"),
  Classification = c("Confirmed (wild)", "Confirmed (wild)", "Discarded"),
  PolioVirusTypes = c("WILD1", "WILD1", NA),
  CountryISO3Code = c("NGA", "NGA", "NGA"),
  Admin0Name = c("NIGERIA", "NIGERIA", "NIGERIA"),
  check.names = FALSE
)
```

The two rows that share `Id = 1` are the same record updated twice; the cleaner
keeps the latest by `LastUpdateDate`.

```{r}
#| label: afp-clean
cases <- clean_afp(afp_raw, cfg = cfg, verbose = FALSE)

cases[, c(
  "epid", "year_onset", "onset_to_notify", "onset_to_stool1",
  "timeliness", "classification_all", "risk_group"
)]
```

The raw classification is decoded into the standard **WPV** / **cVDPV** /
**aVDPV** / **iVDPV** vocabulary, so a single filter works across every stream:

```{r}
#| label: afp-class
cases[grepl("WPV|VDPV", cases$classification_all), c("epid", "classification_all")]
```

Supplying a cleaned district `shape` adds GUID reconciliation and coordinate /
EPID-prefix admin recovery (see the *EPID geography* and *spatial data*
vignettes); without one, `clean_afp()` still runs standalone as above.

## Human lab specimens: `clean_human_spec()`

`clean_human_spec()` is the specimen-level companion to `clean_afp()`: where the
Case table summarises stool 1/2, the LabSpecimen table carries every specimen
with its laboratory results. It derives the same virus classification, a
specimen `adequate` flag, and — since specimens have no onset date — the
**lab-turnaround** intervals (collection → lab → culture → ITD → sequencing) in
place of the onset-based ones.

```{r}
#| label: spec
spec_raw <- data.frame(
  Id = 1:2,
  SpecimenId = c("S-1", "S-2"),
  Epid = c("NIE-BOR-MMC-24-001", "NIE-BOR-JER-24-014"),
  LastUpdateDate = c("2024-03-01", "2024-03-01"),
  DateStoolCollected = c("2024-01-05", "2024-02-09"),
  DateStoolReceivedInLab = c("2024-01-10", "2024-02-14"),
  DateFinalCellCultureResults = c("2024-01-20", NA),
  VirusTypes = c("cVDPV2", NA),
  VdpvClassification = c("Circulating", NA),
  AdequateSpecimen = c("Yes", "No"),
  Admin0Name = c("NIGERIA", "NIGERIA"),
  check.names = FALSE
)

spec <- clean_human_spec(spec_raw, verbose = FALSE)
spec[, c(
  "specimen_id", "classification_all", "adequate",
  "collect_to_lab", "lab_to_culture"
)]
```

## Environmental samples: `clean_es()`

`clean_es()` cleans the EnvSamples table and derives the same virus
classification plus the ES detection flags (`sabin1/2/3`, `npev`, `nvaccine`,
and the fused `ev_detect` "anything detected" flag).

```{r}
#| label: es
es_raw <- data.frame(
  Id = c(1, 1, 2),
  SampleId = c("E-1", "E-1", "E-2"),
  LastUpdateDate = c("2024-02-01", "2024-04-01", "2024-03-01"),
  CollectionDate = c("2024-01-05", "2024-01-05", "2024-02-09"),
  VirusTypes = c("cVDPV2", "cVDPV2", "NPEV"),
  VdpvClassifications = c("Circulating", "Circulating", NA),
  IsNPEV = c(NA, NA, TRUE),
  Admin0Name = c("NIGERIA", "NIGERIA", "CHAD"),
  CountryISO3Code = c("NGA", "NGA", "TCD"),
  check.names = FALSE
)

es <- clean_es(es_raw, verbose = FALSE)
es[, c(
  "sample_id", "year_collection", "classification_all",
  "ev_detect", "npev", "risk_group"
)]
```

## Immunisation campaigns: `clean_sia()`

`clean_sia()` combines the Activity and SubActivity tables into one analytic SIA
table on the sub-activity grain (one row per round × district), joining the
parent campaign on by the sub-activity code and deriving the campaign start
year/month.

```{r}
#| label: sia
activity <- data.frame(
  Id = 1,
  SIASubActivityCode = "S1",
  LastUpdateDate = "2024-03-01",
  VaccineType = "bOPV",
  check.names = FALSE
)
subactivity <- data.frame(
  Id = 10,
  SIASubActivityCode = "S1",
  LastModificationDate = "2024-03-01",
  DateFrom = "2024-03-10",
  Admin0Name = "NIGERIA",
  check.names = FALSE
)

sia <- clean_sia(activity, subactivity, verbose = FALSE)
sia[, c("id", "adm0", "year_start", "month_start", "vaccine_type")]
```

## Poliovirus positives: `clean_virus()`

`clean_virus()` does **not** read a raw viruses table — it *builds* the positives
dataset from the already-cleaned AFP and ES outputs. Every poliovirus-positive
case or sample becomes one harmonised row, tagged by `surveillance_type`, with a
`report_date` (VDPV classification-change date, or notification date for WPV).

```{r}
#| label: virus
positives <- clean_virus(cases = cases, es = es, verbose = FALSE)
positives[, c(
  "epid", "surveillance_type", "measurement", "classification_all", "report_date"
)]
```

Pass `separate_rows = TRUE` to split a co-detection (e.g. `WPV1andcVDPV 2`) into
one row per serotype.

## One call: `run_pipeline()`

`run_pipeline()` runs the per-stream cleaners over a named list of raw tables and
returns the cleaned set, building the virus positives from the cleaned AFP/ES
streams automatically.

```{r}
#| label: pipeline
cleaned <- run_pipeline(
  inputs = list(afp = afp_raw, es = es_raw),
  cfg = cfg
)
names(cleaned)
```

Each element is exactly what the matching `clean_*()` call produces above, so
you can start with `run_pipeline()` and drop down to an individual cleaner
whenever you need its stream-specific arguments (a `shape` for geography
reconciliation, `sites` for ES site validation, and so on).
