---
title: "Data-quality checks"
vignette: >
  %\VignetteIndexEntry{Data-quality checks}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    df-print: default
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    message: false
    warning: false
---

```{r}
#| label: setup
#| include: false
library(polished)
```

The check functions identify duplicates, blank keys, unreconciled administrative
GUIDs, out-of-range values and dates in the wrong order. They use the columns
produced by the cleaners.

There is one function per stream:

| Function | Stream |
|---|---|
| `checks_afp()` | cleaned AFP cases |
| `checks_es()` | cleaned environmental samples |
| `checks_hum_spec()` | cleaned human specimens |
| `checks_sia()` | cleaned SIA campaigns |
| `checks_virus()` | cleaned poliovirus positives |

## Running a check

Each `checks_*()` takes a cleaned table and returns a list: a `summary` tibble
listing each configured check, plus one tibble of flagged rows per check that
found problems. The example below creates a small cleaned AFP table with a few
deliberate issues:

```{r}
#| label: fixture
afp <- tibble::tibble(
  id = 1:5,
  epid = c("NIE-A-1", "NIE-A-1", "NIE-C-3", "NIE-D-4", "NIE-E-5"),
  adm0 = c("NIGERIA", "NIGERIA", "CHAD", "MALI", "NIGER"),
  adm1 = "p", adm2 = "d",
  paralysis_onset_date = c("2024-01-02", "2024-01-02", NA, "2999-01-01", "2024-03-03"),
  year_onset = c(2024L, 2024L, NA, 2999L, 2024L),
  classification_all = c("NPAFP", "NPAFP", "", "NPAFP", "NPAFP"),
  adm1_guid = c("g", "g", NA, "g", "g"),
  adm2_guid = "g",
  latitude = c(9.1, 9.1, 0, 9.1, 9.1),
  longitude = 7.2,
  age_months = c(24, 24, -5, 30, 36),
  notify_to_invest = c(1, 1, -1, 2, 1),
  adequate_stool = c("Yes", "Yes", "No", "Yes", "Yes")
)

res <- checks_afp(afp)
res$summary
```

The `summary` lists each check, its `severity` (`error` / `warning` / `info`,
sorted worst first), and how many rows it flagged. Checks whose required columns
are absent have `status = "not_run"`, `n_flagged = NA`, and the missing column
names in `missing_columns`. A check that ran and found no issues has
`n_flagged = 0`.

Each check with flagged rows has a named table containing the record keys and
the columns needed to review the issue:

```{r}
#| label: detail
names(res)
res$afp_missing_guid
```

## Export to Excel

`write_checks_excel()` turns a check result into one styled `.xlsx` workbook — a
`Summary` tab plus one tab of flagged rows per failing check (navy headers, sized
columns, inferred number formats). There is no versioning; the file is written
straight to `path`.

```{r}
#| label: excel
#| eval: false
write_checks_excel(res, "checks_afp.xlsx")
```

## Automatic check workbooks in the pipeline

When you run the file-based pipeline with an output directory,
`run_pipeline_dir()` writes a `checks_<dataset>.xlsx` workbook next to each
`polished_*` output automatically — so a single call produces both the cleaned
data and its quality report (see the *End-to-end pipeline* article). This step
needs the optional `openxlsx` package; without it the checks are skipped with a
message rather than failing the run.
