---
title: "Downloading POLIS data"
vignette: >
  %\VignetteIndexEntry{Downloading POLIS data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    df-print: default
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    message: false
    warning: false
---

::: callout-note
The download calls in this vignette talk to the live POLIS service and are shown
**unevaluated** — they need a valid `POLIS_API_KEY` and network access. The only
chunk that runs at build time is the table catalogue, which is static package
data. Copy the calls into a session with a key set to try them.
:::

## What `get_polis_data()` is for

POLIS exposes its data through an OData API, but three quirks make a naive OData
client silently lose rows. `polished::get_polis_data()` is a single entry point
that works around all three, caches results on disk, and resumes cleanly if a
pull is interrupted.

The three traps it handles for you:

- **Date filters are year-aligned only.** POLIS honours a `field le YYYY-12-31`
  bound but returns **zero rows** for any sub-year `le`. `get_polis_data()`
  aligns your `min_date` / `max_date` to year boundaries before building a
  filter, so a mid-year range never silently empties the result.
- **`$skip` is rejected and there is no `@odata.nextLink`.** POLIS caps `$top`
  at 2000 rows and refuses the standard OData paging mechanism. The only way past
  2000 rows is *Id-range pagination* — `…&$orderby=Id&$top=2000&$filter=… and Id
  gt <last>` — which the function does for you, one page at a time.
- **The clinical date columns are sparsely populated.** Columns like `CaseDate`
  and `VirusDate` are `NULL` for many historical records, so filtering on them
  drops rows. The function instead filters on each table's *update* column
  (`LastUpdateDate` / `UpdatedDate` / `Start` / `PublishDate`), which probes
  confirm is fully populated.

## Quick start

Set your key once, then pull a table. By default nothing is returned — the data
lands on disk and you read it when you need it.

```{r}
#| label: quick-start
#| eval: false
Sys.setenv(POLIS_API_KEY = "your-key") # or set it in .Renviron

# Pull the immunization (`im`) table; the data is written to the per-user
# cache and nothing is returned
polished::get_polis_data(tables = "im")

# Read the table back from disk when you need it (written under its raw_* name)
cache <- tools::R_user_dir("polished", which = "cache")
im <- readRDS(file.path(cache, "raw_im.rds"))
```

## The table catalogue

`tables` accepts any of the names in `polis_tables_mapping`, the static catalogue
the function ships with. Passing `tables = NULL` (the default) downloads them
all. This is the one chunk in the vignette that runs at build time, because it is
just package data:

```{r}
#| label: catalogue
polished::polis_tables_mapping
```

Each row records the short `table_name` you pass to `tables =`, the OData
`endpoint` it maps to, the `date_field` used for both filtering and the
duplicate-resolution tiebreak, and the `file_stem` the table is written under on
disk. That stem is the `raw_*` name the cleaning pipeline reads (e.g. `case` is
saved as `raw_afp`), so the download and cleaning halves share one naming
convention. An unknown name aborts with the list of valid ones, so a typo fails
fast rather than fetching nothing.

## Choosing what to fetch

Four arguments narrow the pull:

| Argument | Effect |
|---|---|
| `min_date` / `max_date` | the date range, aligned to whole years (default `2000-01-01` to today) |
| `region` | a WHO region filter (`"Global"`, `"AFRO"`, `"EMRO"`, …) |
| `country_code` | an ISO3 code (e.g. `"NGA"`) added as an exact-match clause |

```{r}
#| label: filtering
#| eval: false
# Nigeria only, 2018 through last year, AFRO region
polished::get_polis_data(
  tables = "case",
  min_date = "2018-01-01",
  max_date = "2024-12-31",
  region = "AFRO",
  country_code = "NGA"
)
```

Because date bounds are year-aligned, `min_date = "2018-06-30"` fetches from
`2018-01-01` regardless — the function never returns fewer rows than the year
covering your bound.

## Where the data lives, and how resume works

Every table is fetched into a per-year *part file*, with a tiny metadata
sidecar, before being merged into one canonical file written under the table's
`raw_*` stem:

```
<polis_folder>/
├── raw_im.rds                  # canonical, merged file you read
└── .parts/raw_im/
    ├── year_2023.rds           # one part per calendar year
    ├── year_2023.meta.rds      # row count + Id range sidecar
    └── year_2024.rds
```

The part files **are the resume marker**. Each year is an independent Id-range
walk that flushes to its part after every 2000-row batch, so if a run dies
mid-pull — a dropped connection, a Ctrl-C, an OOM — the next call picks up at `Id
gt max(Id)` for each year. Worst-case lost work is the single in-flight batch.

Files written by older versions under the bare table name (e.g. `case.rds`,
`.parts/case/`) are renamed to their `raw_*` stem in place on the next run, so an
existing cache is reused rather than re-downloaded.

`polis_folder` defaults to `tools::R_user_dir("polished", which = "cache")`, the
standard per-user cache location, so incremental updates persist across sessions
without you choosing a path. Pass an explicit folder to keep data alongside a
project instead:

```{r}
#| label: project-folder
#| eval: false
polished::get_polis_data(tables = "im", polis_folder = "data/polis")
```

### Incremental updates

Re-running the same call is the update path. The function asks POLIS for the
current row count, compares it against what the part files already hold, and:

- if the cache already covers the count, it **skips the pull entirely** (just
  re-merging parts to the canonical file); otherwise
- it walks each year from its last cached `Id` forward, fetching only the new
  rows.

So a daily cron that calls `get_polis_data(tables = "case")` downloads the full
history once and only the delta thereafter.

To force a clean re-pull instead of resuming, pass `force = TRUE` (this deletes
the table's parts and canonical file first).

By default (`prune_parts = TRUE`) the per-year parts are deleted once the
canonical file is written and verified; the next run rebuilds them from the
canonical, which keeps the parts free of stale cross-year duplicates and keeps
the "already up to date" short-circuit honest. Incremental resume still works.
Pass `prune_parts = FALSE` to keep the parts on disk for the fastest possible
resume (no re-split next run).

## Going faster with parallel workers

Each calendar year is an independent walk, so years can be fetched concurrently.
`workers > 1` dispatches them across a `parallel::makePSOCKcluster()` — the same
transport on Windows, macOS, and Linux — while a single live progress bar polls
the part files so you watch rows accumulate across workers in real time:

```{r}
#| label: parallel
#| eval: false
polished::get_polis_data(
  tables = "virus",
  workers = parallel::detectCores() - 1L
)
```

::: callout-important
PSOCK workers start fresh R sessions and load `polished` with `library()`, so the
package must be **installed** for parallel mode — a `devtools::load_all()` session
is not enough. With `workers = 1L` (the default) a single sequential loop drives
the bar per batch and has no such requirement.
:::

## What you get back

Nothing — and that's the point. `get_polis_data()` is called purely for its
**side effect** of writing each table to disk, and returns `NULL` invisibly. A
single POLIS table can run to millions of rows, so returning that into memory on
every call would be a footgun; instead the data stays on disk and you read back
only the table you need, only when you need it.

Each table lands at `<polis_folder>/<file_stem>.<ext>` (the `raw_*` name from the
catalogue). Read one back with the matching reader for your `output_format`:

```{r}
#| label: read-back
#| eval: false
# default per-user cache location
cache <- tools::R_user_dir("polished", which = "cache")

case <- readRDS(file.path(cache, "raw_afp.rds")) # `case` is saved as raw_afp
# or, for other output formats:
# arrow::read_parquet(file.path(cache, "raw_afp.parquet"))
```

Because the value is invisible, a bare `get_polis_data(...)` at the console
prints nothing — assign the call if you want the paths.

### Output formats

`output_format` controls how the canonical file is written: `"rds"` (default),
`"rda"`, `"csv"`, `"parquet"`, or `"qs2"`. The `parquet` and `qs2` formats need
the `arrow` and `qs2` packages respectively.

```{r}
#| label: formats
#| eval: false
polished::get_polis_data(tables = "case", output_format = "parquet")
```

## Completeness verification

POLIS occasionally truncates a query under load and returns a partial page even
when more rows exist. With `auto_refetch = TRUE` (the default), after each table
finishes the function:

1. uses the metadata sidecars to detect a gap cheaply (row-count and Id-range
   mismatch against POLIS's reported `@odata.count`);
2. only if a gap is found, issues a lightweight `$select=Id` probe to list the
   canonical Id set; and
3. refetches any missing Ids via `Id in (...)` chunks and merges them in,
   de-duplicating by `Id` keeping the latest update date.

Set `auto_refetch = FALSE` to trust whatever is on disk and skip the check — for
example in the smoke-test path, or when you know the pull completed cleanly and
want to save the verification round-trip.

## Other options worth knowing

| Argument | What it does |
|---|---|
| `keep_archives` | when `> 0`, also writes a timestamped copy under `archive/` on each save and prunes older copies beyond this many |
| `prune_parts` | when `TRUE` (default), deletes the `.parts/` resume cache after the canonical is written and verified; rebuilt from the canonical next run |
| `log_file` | path to a per-batch `.rds` log of what was fetched, when, and how many rows |
| `quiet` | suppresses headers, progress bars, and the info alerts |
| `polis_api_key` | the key; defaults to `Sys.getenv("POLIS_API_KEY")` |

## From download to clean

`get_polis_data()` gets you a faithful local copy of POLIS; the cleaners take it
from there. A typical flow pulls a table, then recovers any missing administrative
geography from the EPID:

```{r}
#| label: download-then-clean
#| eval: false
polished::get_polis_data(tables = "case")

cache <- tools::R_user_dir("polished", which = "cache")
cases <- readRDS(file.path(cache, "raw_afp.rds")) # `case` is saved as raw_afp

cleaned <- polished::impute_geo_from_epid(cases)
cleaned$qa # what was filled, and what was left unresolved
```

See the *Recovering geography from EPIDs* vignette for that second step in
detail, the *End-to-end pipeline* article to go from a folder of `raw_*` files
straight to `polished_*` outputs, and `?get_polis_data` for the complete
argument reference.
