---
title: "Downloading POLIS data"
vignette: >
  %\VignetteIndexEntry{Downloading POLIS data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    df-print: default
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    message: false
    warning: false
---

::: callout-note
The download calls in this vignette talk to the live POLIS service and are shown
**unevaluated** — they need a valid `POLIS_API_KEY` and network access. The only
chunk that runs at build time is the table catalogue, which is static package
data. Copy the calls into a session with a key set to try them.
:::

## What `get_polis_data()` is for

POLIS exposes its data through an OData API, but three quirks make a naive OData
client silently lose rows. `polished::get_polis_data()` is a single entry point
that works around all three, caches results on disk, and resumes cleanly if a
pull is interrupted.

The three traps it handles for you:

- **Date filters are year-aligned only.** POLIS honours a `field le YYYY-12-31`
  bound but returns **zero rows** for any sub-year `le`. `get_polis_data()`
  aligns your `min_date` / `max_date` to year boundaries before building a
  filter, so a mid-year range never silently empties the result.
- **`$skip` is rejected and there is no `@odata.nextLink`.** POLIS caps `$top`
  at 2000 rows and refuses the standard OData paging mechanism. The only way past
  2000 rows is *Id-range pagination* — `…&$orderby=Id&$top=2000&$filter=… and Id
  gt <last>` — which the function does for you, one page at a time.
- **The clinical date columns are sparsely populated.** Columns like `CaseDate`
  and `VirusDate` are `NULL` for many historical records, so filtering on them
  drops rows. The function instead filters on each table's *update* column
  (`LastUpdateDate` / `UpdatedDate` / `Start` / `PublishDate`), which probes
  confirm is fully populated.

## Quick start

Set your key once, then pull a table. By default nothing is returned — the data
lands on disk and you read it when you need it.

```{r}
#| label: quick-start
#| eval: false
Sys.setenv(POLIS_API_KEY = "your-key") # or set it in .Renviron

# Pull the immunization (`im`) table; the data is written to the per-user
# cache and nothing is returned
polished::get_polis_data(tables = "im")

# Read the table back from disk when you need it (written under its raw_* name)
cache <- tools::R_user_dir("polished", which = "cache")
im <- readRDS(file.path(cache, "raw_im.rds"))
```

## The table catalogue

`tables` accepts any of the names in `polis_tables_mapping`, the static catalogue
the function ships with. Passing `tables = NULL` (the default) downloads them
all. This is the one chunk in the vignette that runs at build time, because it is
just package data:

```{r}
#| label: catalogue
polished::polis_tables_mapping
```

Each row records the short `table_name` you pass to `tables =`, the OData
`endpoint` it maps to, the `date_field` used for both filtering and the
duplicate-resolution tiebreak, and the `file_stem` the table is written under on
disk. That stem is the `raw_*` name the cleaning pipeline reads (e.g. `case` is
saved as `raw_afp`), so the download and cleaning halves share one naming
convention. An unknown name aborts with the list of valid ones, so a typo fails
fast rather than fetching nothing.

## Choosing what to fetch

Four arguments narrow the pull:

| Argument | Effect |
|---|---|
| `min_date` / `max_date` | the date range, aligned to whole years (default `2000-01-01` to today) |
| `region` | a WHO region filter (`"Global"`, `"AFRO"`, `"EMRO"`, …) |
| `country_code` | an ISO3 code (e.g. `"NGA"`) added as an exact-match clause |

```{r}
#| label: filtering
#| eval: false
# Nigeria only, 2018 through last year, AFRO region
polished::get_polis_data(
  tables = "case",
  min_date = "2018-01-01",
  max_date = "2024-12-31",
  region = "AFRO",
  country_code = "NGA"
)
```

Because date bounds are year-aligned, `min_date = "2018-06-30"` fetches from
`2018-01-01` regardless — the function never returns fewer rows than the year
covering your bound.

## Where the data lives, and how resume works

Each batch is written as an immutable page and committed in a journal. Once a
year finishes, its pages are compacted into one part. The final merge writes the
canonical file under the table's `raw_*` stem:

```
<polis_folder>/
├── raw_afp.rds
├── .manifests/raw_afp.rds.rds   # scope, revisions and completion state
└── .parts/raw_afp/
    ├── year_2023.rds
    ├── year_2023.meta.rds      # progress without rereading the data
    └── year_2024.rds.pages/
        ├── state.rds          # committed pages and last Id
        └── page_00000001.rds
```

An interrupted run resumes after the last committed page. A page whose journal
commit failed is fetched again. The final merge needs memory for the complete
table. A failed refresh leaves the previous canonical file in place.

Files with older names such as `case.rds` are renamed to `raw_afp.rds`. Legacy
caches without scope metadata require one fresh download because the country,
region and date filters used to create them are unknown.

`polis_folder` defaults to `tools::R_user_dir("polished", which = "cache")`, the
standard per-user cache location, so incremental updates persist across sessions
without you choosing a path. Pass an explicit folder to keep data alongside a
project instead:

```{r}
#| label: project-folder
#| eval: false
polished::get_polis_data(tables = "im", polis_folder = "data/polis")
```

### Incremental updates

Re-running a call checks the saved scope before reusing its data. Country,
region or year-range changes start a fresh pull. For tables with
`LastUpdateDate` or `UpdatedDate`, the downloader compares the full scoped Id
and revision list, fetches changed rows, and removes deleted records. This
assumes the service advances its revision timestamp on edits. Equal row counts
do not establish freshness.

An unchanged snapshot needs no canonical read or partition rebuilding. Tables
filtered by event dates (`lqas` and `im`) are fetched again because those dates
do not identify edits. Population snapshots have a one-day expiry by default;
use `reference_refresh_days = 0` to refresh them on every call.

`force = TRUE` starts a full pull while retaining the old canonical until its
replacement succeeds. `auto_refetch = FALSE` explicitly trusts completed
snapshots with matching scope until forced.

By default, `prune_parts = TRUE` removes completed parts. Use `FALSE` to retain
them for inspection. Revision changes discard retained parts so they cannot
restore stale records on a later run.

## Going faster with parallel workers

Each calendar year is an independent walk, so years can be fetched concurrently.
`workers > 1` dispatches them across a `parallel::makePSOCKcluster()` — the same
transport on Windows, macOS, and Linux — while a single live progress bar polls
the part files so you watch rows accumulate across workers in real time:

```{r}
#| label: parallel
#| eval: false
polished::get_polis_data(
  tables = "virus",
  workers = parallel::detectCores() - 1L
)
```

::: callout-important
PSOCK workers start fresh R sessions and load `polished` with `library()`, so the
package must be **installed** for parallel mode — a `devtools::load_all()` session
is not enough. With `workers = 1L` (the default) a single sequential loop drives
the bar per batch and has no such requirement.
:::

## What you get back

Nothing — and that's the point. `get_polis_data()` is called purely for its
**side effect** of writing each table to disk, and returns `NULL` invisibly. A
single POLIS table can run to millions of rows, so returning that into memory on
every call would be a footgun; instead the data stays on disk and you read back
only the table you need, only when you need it.

Each table lands at `<polis_folder>/<file_stem>.<ext>` (the `raw_*` name from the
catalogue). Read one back with the matching reader for your `output_format`:

```{r}
#| label: read-back
#| eval: false
# default per-user cache location
cache <- tools::R_user_dir("polished", which = "cache")

case <- readRDS(file.path(cache, "raw_afp.rds")) # `case` is saved as raw_afp
# or, for other output formats:
# arrow::read_parquet(file.path(cache, "raw_afp.parquet"))
```

Because the value is invisible, a bare `get_polis_data(...)` at the console
prints nothing — assign the call if you want the paths.

### Output formats

`output_format` controls how the canonical file is written: `"rds"` (default),
`"rda"`, `"csv"`, `"parquet"`, or `"qs2"`. The `parquet` and `qs2` formats need
the `arrow` and `qs2` packages respectively.

```{r}
#| label: formats
#| eval: false
polished::get_polis_data(tables = "case", output_format = "parquet")
```

## Completeness verification

POLIS occasionally truncates a query under load and returns a partial page even
when more rows exist. With `auto_refetch = TRUE` (the default), after each table
finishes the function:

1. uses the metadata sidecars to detect a gap cheaply (row-count and Id-range
   mismatch against POLIS's reported `@odata.count`);
2. only if a gap is found, issues a lightweight `$select=Id` probe to list the
   canonical Id set; and
3. refetches any missing Ids via `Id in (...)` chunks and merges them in,
   de-duplicating by `Id` keeping the latest update date.

Set `auto_refetch = FALSE` to trust whatever is on disk and skip the check — for
example in the smoke-test path, or when you know the pull completed cleanly and
want to save the verification round-trip.

## Other options worth knowing

| Argument | What it does |
|---|---|
| `keep_archives` | when `> 0`, also writes a timestamped copy under `archive/` on each save and prunes older copies beyond this many |
| `prune_parts` | when `TRUE` (default), deletes the `.parts/` resume cache after the canonical is written and verified; rebuilt from the canonical next run |
| `log_file` | path to a per-batch `.rds` log of what was fetched, when, and how many rows |
| `quiet` | suppresses headers, progress bars, and the info alerts |
| `polis_api_key` | the key; defaults to `Sys.getenv("POLIS_API_KEY")` |

## From download to clean

`get_polis_data()` gets you a faithful local copy of POLIS; the cleaners take it
from there. A typical flow pulls a table, then recovers any missing administrative
geography from the EPID:

```{r}
#| label: download-then-clean
#| eval: false
polished::get_polis_data(tables = "case")

cache <- tools::R_user_dir("polished", which = "cache")
cases <- readRDS(file.path(cache, "raw_afp.rds")) # `case` is saved as raw_afp

cleaned <- polished::impute_geo_from_epid(cases)
cleaned$qa # what was filled, and what was left unresolved
```

See the *Recovering geography from EPIDs* vignette for that second step in
detail, the *End-to-end pipeline* article to go from a folder of `raw_*` files
straight to `polished_*` outputs, and `?get_polis_data` for the complete
argument reference.
