---
title: "Recovering geography from EPIDs"
vignette: >
  %\VignetteIndexEntry{Recovering geography from EPIDs}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    df-print: paged
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    message: false
    warning: false
---

```{r}
#| label: setup
library(polished)
library(tibble)
library(dplyr)
```

::: callout-note
Every data frame in this vignette is **synthetic** — fabricated EPIDs and
place names, never real case data. The geography names are illustrative.
:::

## What this module does

An **EPID** is a hierarchical case identifier, conventionally shaped as
`COUNTRY-PROVINCE-DISTRICT-YEAR-SERIAL` (e.g. `NIE-BOR-MMC-24-001`). The leading
three characters are a country code; the next segments encode province/district
abbreviations. When a line-list arrives with **missing admin names or GUIDs**
but a populated EPID, the EPID is a recovery key for geography.

The module separates two concerns:

- **Parsing (A)** — pure functions that pull *codes* out of an EPID string.
- **Imputation (B)** — a provenance-stamped cascade that fills *names and GUIDs*
  from sibling records or an external reference.

Three principles hold throughout:

1. **Never overwrite** a value that is already present.
2. **Never fabricate** on ambiguity — if two candidates disagree, leave `NA` and
   flag it.
3. **Always stamp provenance** — every filled cell records *how* it was filled.

### The public surface

These eight functions are exported and documented here:

| Function | Role |
|---|---|
| `epid_split()` | parse an EPID into its segments |
| `epid_country_code()` | extract the country code |
| `epid_prefix()` | the geographic prefix used for matching |
| `epid_strip_contact()` | split a contact marker off the base EPID |
| `build_admin_ref()` | most-recent value per EPID |
| `build_prefix_ref()` | unique value per (prefix, year) |
| `resolve_epid_country()` | code → country name / ISO3 |
| `impute_geo_from_epid()` | the orchestrator running the full cascade |

Everything else in the module (`.epid_normalise()`, `.epid_prefix_match()`,
`.epid_fill_target()`, …) is an **internal helper** — `.`-prefixed, not
exported, and not part of the public API.

## Parsing EPIDs (capability A)

### `epid_split()` — segments

`epid_split()` never errors, whatever you throw at it: short EPIDs pad on the
right with `NA`, blank/`NA` rows come back all-`NA`.

```{r}
#| label: epid-split
epid_split(c(
  "NIE-BOR-MMC-24-001", # canonical
  "AGO-LUA", # short — trailing parts NA
  "nie-yob-dam-24-014", # lowercase still splits
  "  ", # blank -> all NA
  NA # NA -> all NA
))
```

By default extra segments are dropped. Use `extra = "merge"` to keep an
unexpected tail (e.g. a free-text suffix) in the last column:

```{r}
#| label: epid-split-merge
epid_split("NIE-BOR-MMC-24-001-RECHECK", extra = "merge")
```

### `epid_country_code()` — the country code

The code is the first run of three word-characters, so it tolerates leading
whitespace and mixed case. It works equally on a local abbreviation (`NIE`) or
an ISO3 (`NGA`) — resolving *which* is the job of `resolve_epid_country()`.

```{r}
#| label: epid-country-code
epid_country_code(c("NIE-BOR-1", "  nga-bor-1", "AGO-LUA", NA))

# `upper = FALSE` keeps the original case
epid_country_code("ago-lua-01", upper = FALSE)
```

### `epid_prefix()` — the matching stem

The prefix is the leading slice of the *normalised* EPID (trimmed,
whitespace-collapsed, upper-cased). The default length of 11 captures
country+province+district for a `CCC-PPP-DDD` layout.

```{r}
#| label: epid-prefix
epid_prefix(c("NIE-BOR-MMC-24-001", "nie-bor-mmc-24-002", "  ", NA))
```

### `epid_strip_contact()` — contacts vs cases

Contact records reuse a case EPID with a trailing marker (`C`, `CC`, `HC`, or
`C` + digits). Stripping it lets a contact **collapse onto its parent case** so
it inherits the case's geography.

```{r}
#| label: epid-strip-contact
epid_strip_contact(c(
  "NIE-BOR-MMC-24-001", # a case — no marker
  "NIE-BOR-MMC-24-001CC", # community contact
  "NIE-BOR-MMC-24-001-HC", # health-facility contact
  "NIE-BOR-MMC-24-001C2" # numbered contact
))
```

Note all four share the same `epid_base` — which is exactly why the cascade can
fill a contact's missing district from its case.

## Building references (capability B inputs)

### `build_admin_ref()` — most-recent per EPID

For each EPID, `build_admin_ref()` keeps the **most recent non-blank** value of a
column. Older and missing values are ignored.

```{r}
#| label: build-admin-ref
case_history <- tibble(
  epid = c("NIE-BOR-MMC-24-001", "NIE-BOR-MMC-24-001", "NIE-BOR-MMC-24-001"),
  year = c(2022, 2024, 2023),
  district = c("MAIDUGURI (OLD)", NA, "MAIDUGURI")
)
build_admin_ref(case_history, "district")
```

The 2024 row is blank, so the most recent *non-blank* value (2023's
`"MAIDUGURI"`) wins — not the stale 2022 spelling.

### `build_prefix_ref()` — unique per (prefix, year)

`build_prefix_ref()` is deliberately conservative: it only returns a value when
the prefix maps to **exactly one** district that year. `n_candidates` records how
many distinct values competed, so you can see where geography is genuinely
ambiguous.

```{r}
#| label: build-prefix-ref
prefix_cases <- tibble(
  epid = c(
    "NIE-BOR-MMC-24-001", "NIE-BOR-MMC-24-050", # both Maiduguri
    "NIE-YOB-DAM-24-001", "NIE-YOB-DAM-24-002" # disagree!
  ),
  year = c(2024, 2024, 2024, 2024),
  district = c("MAIDUGURI", "MAIDUGURI", "DAMATURU", "POTISKUM")
)
build_prefix_ref(prefix_cases, "district", prefix_length = 11)
```

The `NIE-BOR-MMC` prefix resolves cleanly; `NIE-YOB-DAM` has two rival districts
(`n_candidates = 2`) so its value is `NA` — the module would rather leave a gap
than guess.

## Resolving the country (capability A → name)

`resolve_epid_country()` maps the country code to a name + ISO3 via a crosswalk
you supply. It matches on **either** the code or the ISO3 column, and refuses to
guess when a code maps to more than one country.

```{r}
#| label: country-crosswalk
country_ref <- tibble(
  code = c("NIE", "AGO", "TCD", "CMR"),
  name = c("NIGERIA", "ANGOLA", "CHAD", "CAMEROON"),
  iso3 = c("NGA", "AGO", "TCD", "CMR")
)

# Matches a local abbreviation and an ISO3 alike
resolve_epid_country(c("NIE-BOR-1", "NGA-BOR-1", "AGO-LUA-1"), ref = country_ref)
```

With no crosswalk, you still get the raw code back (with `resolved = FALSE`) —
never a fabricated name:

```{r}
#| label: country-no-ref
resolve_epid_country("NIE-BOR-1")
```

A code that maps to two different names is flagged `ambiguous` and left `NA`:

```{r}
#| label: country-ambiguous
clashing_ref <- tibble(
  code = c("CHA", "CHA"),
  name = c("CHAD", "CENTRAL AFRICAN REPUBLIC"),
  iso3 = c("TCD", "CAF")
)
resolve_epid_country("CHA-XXX-1", ref = clashing_ref)
```

## The imputation cascade — `impute_geo_from_epid()`

The orchestrator fills each admin level (Admin0, then Admin1, then Admin2) by
trying strategies in order, stamping a `<col>_source` provenance factor as it
goes. The strategies:

1. **`self_ref`** — the same exact EPID elsewhere in the data.
2. **`prefix_match`** — a unique value among records sharing the prefix within
   `year_window` years, with a **parent tie-break** before declaring ambiguity.
3. **`reference`** — an external table keyed on `epid` or `prefix`.
4. **`country_prefix`** — Admin0 only, via the country crosswalk.

### `self_ref` — same EPID, including contacts

Here a community contact (`…001CC`) is missing its district; its parent case
carries it. Because contacts strip to the same base EPID, `self_ref` fills it.

```{r}
#| label: cascade-self-ref
contacts <- tibble(
  epid = c("NIE-BOR-MMC-24-001", "NIE-BOR-MMC-24-001CC"),
  yronset = c(2024, 2024),
  place.admin.2 = c("MAIDUGURI", NA)
)
self_ref_out <- impute_geo_from_epid(
  contacts,
  admin0_var = NULL, admin1_var = NULL, guid_vars = NULL,
  strategies = "self_ref",
  verbose = FALSE
)
self_ref_out$data |>
  select(epid, place.admin.2, place.admin.2_source)
```

### `prefix_match` — siblings within a year window

Different serials under the same `CCC-PPP-DDD` prefix share a district. With
`year_window = 1`, a case can borrow from a neighbour in an adjacent year.

```{r}
#| label: cascade-prefix
siblings <- tibble(
  epid = c(
    "NIE-BOR-JER-23-014", # known, prior year
    "NIE-BOR-JER-24-002" # missing district
  ),
  yronset = c(2023, 2024),
  place.admin.2 = c("JERE", NA)
)
impute_geo_from_epid(
  siblings,
  admin0_var = NULL, admin1_var = NULL, guid_vars = NULL,
  strategies = "prefix_match", prefix_length = 11, year_window = 1,
  verbose = FALSE
)$data |>
  select(epid, place.admin.2, place.admin.2_source)
```

### Ambiguity is left unresolved, not guessed

When one prefix maps to two districts in the same year and the parent province
can't break the tie, the cell stays `NA`, the run counts it as ambiguous, and
the provenance is `unresolved`.

```{r}
#| label: cascade-ambiguous
ambiguous <- tibble(
  epid = c(
    "NIE-YOB-DAM-24-001", "NIE-YOB-DAM-24-002", "NIE-YOB-DAM-24-003"
  ),
  yronset = c(2024, 2024, 2024),
  place.admin.1 = c("YOBE", "YOBE", "YOBE"),
  place.admin.2 = c("DAMATURU", "POTISKUM", NA)
)
amb_out <- impute_geo_from_epid(
  ambiguous,
  admin0_var = NULL, guid_vars = NULL,
  strategies = "prefix_match", prefix_length = 11,
  verbose = FALSE
)
amb_out$data |> select(epid, place.admin.2, place.admin.2_source)
amb_out$qa |> select(column, n_missing_before, n_ambiguous, n_unresolved)
```

### The parent tie-break

If two districts share a prefix but sit in **different provinces**, a row whose
province *is* known resolves cleanly to the matching district.

```{r}
#| label: cascade-tiebreak
tiebreak <- tibble(
  epid = c(
    "NIE-XXX-AAA-24-001", "NIE-XXX-AAA-24-002", "NIE-XXX-AAA-24-003"
  ),
  yronset = c(2024, 2024, 2024),
  place.admin.1 = c("BORNO", "YOBE", "BORNO"), # row 3's province is known
  place.admin.2 = c("MAIDUGURI", "DAMATURU", NA)
)
impute_geo_from_epid(
  tiebreak,
  admin0_var = NULL, guid_vars = NULL,
  strategies = "prefix_match", prefix_length = 11,
  verbose = FALSE
)$data |>
  select(epid, place.admin.1, place.admin.2, place.admin.2_source)
```

Row 3 is in `BORNO`, so it takes `MAIDUGURI`, not `DAMATURU`.

### `reference` — a dataset with no names of its own

Some extracts carry only EPIDs. An external `reference` keyed on `epid` (or
`prefix`) fills them. Here there are no in-data names to borrow at all.

```{r}
#| label: cascade-reference
nameless <- tibble(
  epid = c("NIE-BOR-MMC-24-001", "AGO-LUA-CAC-24-001"),
  place.admin.2 = c(NA_character_, NA_character_)
)
district_lookup <- tibble(
  epid = c("NIE-BOR-MMC-24-001", "AGO-LUA-CAC-24-001"),
  place.admin.2 = c("MAIDUGURI", "CACUACO")
)
impute_geo_from_epid(
  nameless,
  admin0_var = NULL, admin1_var = NULL, guid_vars = NULL,
  reference = district_lookup, strategies = "reference",
  verbose = FALSE
)$data |>
  select(epid, place.admin.2, place.admin.2_source)
```

### `country_prefix` — Admin0 from the crosswalk

Admin0 is best recovered from the country code. With a crosswalk, a missing
country name is filled from the leading code.

```{r}
#| label: cascade-country
no_country <- tibble(
  epid = c("NIE-BOR-MMC-24-001", "AGO-LUA-CAC-24-001"),
  place.admin.0 = c(NA_character_, NA_character_)
)
impute_geo_from_epid(
  no_country,
  admin1_var = NULL, admin2_var = NULL, guid_vars = NULL,
  country_ref = country_ref, strategies = "country_prefix",
  canonicalise = FALSE,
  verbose = FALSE
)$data |>
  select(epid, place.admin.0, place.admin.0_source)
```

### Filling GUIDs too

GUID columns ride the same cascade. Point `guid_vars` at them, keyed by level:

```{r}
#| label: cascade-guid
with_guids <- tibble(
  epid = c("NIE-BOR-MMC-24-001", "NIE-BOR-MMC-24-002"),
  yronset = c(2024, 2024),
  place.admin.2 = c("MAIDUGURI", "MAIDUGURI"),
  Admin2GUID = c("guid-maiduguri", NA)
)
impute_geo_from_epid(
  with_guids,
  admin0_var = NULL, admin1_var = NULL,
  guid_vars = c(adm2 = "Admin2GUID"),
  strategies = "prefix_match", prefix_length = 11,
  verbose = FALSE
)$data |>
  select(epid, Admin2GUID, Admin2GUID_source)
```

## Putting it together: a messy line-list

A realistic extract mixes all of the above — complete rows, contacts, sibling
gaps, a lowercase EPID, a missing country, a genuinely ambiguous prefix, and a
row nothing can rescue. Run the full cascade with both reference tables and a
one-year window, and let the cli summary report what happened.

```{r}
#| label: messy-linelist
afp_linelist <- tibble(
  epid = c(
    "NIE-BOR-MMC-24-001", # complete
    "NIE-BOR-MMC-24-002", # district gap -> prefix_match
    "NIE-BOR-MMC-24-001CC", # contact -> self_ref off its case
    "nie-bor-jer-24-014", # lowercase, complete
    "NIE-BOR-JER-24-021", # district gap -> prefix_match
    "AGO-LUA-CAC-24-001", # country gap -> country_prefix
    "NIE-YOB-DAM-24-001", # rival district A
    "NIE-YOB-DAM-24-002", # rival district B
    "NIE-YOB-DAM-24-009", # ambiguous -> unresolved
    "NIE-KAN-XYZ-24-001" # unique, no help anywhere -> unresolved
  ),
  yronset = c(2024, 2024, 2024, 2024, 2024, 2024, 2024, 2024, 2024, 2024),
  place.admin.0 = c(
    "NIGERIA", "NIGERIA", "NIGERIA", "NIGERIA", "NIGERIA",
    NA, "NIGERIA", "NIGERIA", "NIGERIA", "NIGERIA"
  ),
  place.admin.1 = c(
    "BORNO", "BORNO", "BORNO", "BORNO", "BORNO",
    "LUANDA", "YOBE", "YOBE", "YOBE", "KANO"
  ),
  place.admin.2 = c(
    "MAIDUGURI", NA, NA, "JERE", NA,
    "CACUACO", "DAMATURU", "POTISKUM", NA, NA
  ),
  Admin2GUID = c(
    "guid-maiduguri", NA, NA, "guid-jere", NA,
    "guid-cacuaco", "guid-damaturu", "guid-potiskum", NA, NA
  )
)

reference_districts <- tibble(
  prefix = c("AGO-LUA-CAC"),
  place.admin.2 = c("CACUACO")
)

result <- impute_geo_from_epid(
  afp_linelist,
  reference = reference_districts,
  country_ref = country_ref,
  guid_vars = c(adm2 = "Admin2GUID"),
  prefix_length = 11,
  year_window = 1
)
```

The returned `data` carries every original column plus one provenance factor per
filled level:

```{r}
#| label: messy-data
result$data |>
  select(
    epid, place.admin.0, place.admin.0_source,
    place.admin.2, place.admin.2_source
  )
```

The `qa` tibble is the audit trail — and its counts always reconcile
(`n_missing_before == filled + n_unresolved`):

```{r}
#| label: messy-qa
result$qa

# reconciliation check
result$qa |>
  mutate(
    filled = n_filled_self_ref + n_filled_prefix_match +
      n_filled_reference + n_filled_country_prefix,
    reconciles = n_missing_before == filled + n_unresolved
  ) |>
  select(column, n_missing_before, filled, n_unresolved, reconciles)
```

`meta` records exactly how the run was configured:

```{r}
#| label: messy-meta
result$meta
```

## Idempotency

Running the cleaner on its own output changes nothing — the second pass sees the
filled values as `original` and has nothing left to do.

```{r}
#| label: idempotent
second_pass <- impute_geo_from_epid(
  result$data,
  reference = reference_districts,
  country_ref = country_ref,
  guid_vars = c(adm2 = "Admin2GUID"),
  prefix_length = 11,
  year_window = 1,
  verbose = FALSE
)
identical(second_pass$data$place.admin.2, result$data$place.admin.2)
```

## Edge cases, in one place

- **Malformed / short / `NA` / blank / lowercase / whitespace EPIDs** are handled
  without error — parsers return `NA`, the cascade skips them.
- **Contacts** (`CC` / `HC` / `C2` suffixes) collapse onto their case EPID for
  matching.
- **Ambiguous prefixes** are never guessed — they stay `NA`, counted in
  `n_ambiguous`, labelled `unresolved`.
- **Present values are never overwritten**; provenance for them is `original`.
- **No crosswalk, no reference, no siblings** → the cell is honestly
  `unresolved`, and the cli summary tells you how many such gaps remain.

```{r}
#| label: edge-recap
messy_epids <- tibble(
  epid = c("nie-bor-mmc-24-001", "  ", NA, "AB", "NIE-BOR-MMC-24-002CC"),
  yronset = c(2024, 2024, 2024, 2024, 2024),
  place.admin.2 = c("MAIDUGURI", NA, NA, NA, NA)
)
impute_geo_from_epid(
  messy_epids,
  admin0_var = NULL, admin1_var = NULL, guid_vars = NULL,
  verbose = FALSE
)$data |>
  select(epid, place.admin.2, place.admin.2_source)
```
