---
title: "Scaffolding a pipeline project"
vignette: >
  %\VignetteIndexEntry{Scaffolding a pipeline project}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    df-print: default
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    message: false
    warning: false
---

```{r}
#| label: setup
#| include: false
library(polished)
```

A real POLIS analysis is more than one `run_pipeline()` call: it has raw
downloads, processed shapefiles, WorldPop rasters, a population reconciliation,
and the cleaned `polished_*` outputs — each with a conventional home on disk, and
a `cfg` that ties them together. `init_polis_pipeline()` stamps out that whole
project in one call: the directory layout, a wired `.Rprofile`, a `.gitignore`,
and runnable `2a_download_data.R` / `2b_process_data.R` starter scripts.

It is the heavier sibling of `init_polis_project()`:

| | `init_polis_project()` | `init_polis_pipeline()` |
|---|---|---|
| Layout | generic `raw/processed/validation/cache/logs` zones | domain-numbered `01_data` (shapefiles / population / polis / vaccination), `02_scripts`, `03_outputs` |
| Writes a `.Rprofile`? | no | yes — the `cfg` manifest |
| Writes starter scripts? | no | yes — download + process |
| Use it when | you just need tidy zones for one stream | you want the full reproducible download → clean pipeline |

## One call

```{r}
#| label: scaffold
root <- file.path(tempdir(), "polio_pipeline")

init_polis_pipeline(
  root,
  regions    = c("EMRO", "AFRO"), # WHO regions the pipeline is scoped to
  start_year = 2015,
  pop_years  = 2010:2027,
  pop_source = "reconciled"       # "reconciled" | "polis" | "worldpop"
)
```

## The layout it produces

```{r}
#| label: tree
cat(sub(root, "<project>", list.dirs(root), fixed = TRUE), sep = "\n")
```

Each `01_data` domain has a `raw/` (inputs) and `processed/` (derived) half:

| Domain | Holds | Produced by |
|---|---|---|
| `1a_shapefiles` | WHO polio GDB layers → `spatial_global_adm{0,1,2}.qs2` | `process_spatial()` (2b) |
| `1b_population/worldpop` | annual rasters → extracted adm2 tables | 2a download + 2b extract |
| `1b_population/polis_pop` | `raw_population.qs2` → cleaned denominators | `get_polis_data()` + `clean_pop()` |
| `1c_polis` | `raw_*.qs2` → `polished_*.qs2` (+ `checks/`, `cache/`) | `get_polis_data()` + `run_pipeline()` |
| `1d_vaccination` | coverage rasters (optional downstream models) | your own scripts |

Pass `domains =` to scaffold a subset (e.g. `c("shapefiles", "polis")`).

## The wired `.Rprofile`

The generated `.Rprofile` is the project **manifest** — the data-directory roots
are defined once as the single source of truth, and `cfg` is built from them, so
the scripts and the cleaners share exactly the same paths:

```{r}
#| label: rprofile
cat(readLines(file.path(root, ".Rprofile")), sep = "\n")
```

The four knobs you passed to `init_polis_pipeline()` are interpolated into the
`polis_config()` call, so editing them later is just editing the `.Rprofile`:

- **`regions`** — scopes the surveillance streams (AFP/ES/SIA/…). Population is
  *foundational* and stays global regardless.
- **`start_year`** — earliest onset/collection year retained.
- **`pop_years`** — the population/WorldPop window.
- **`pop_source`** — which value `clean_pop()` chooses as `<age>_pop`:
  `"reconciled"` (POLIS, with WorldPop fallback on missing/implausible values),
  `"polis"` (POLIS only), or `"worldpop"` (WorldPop only). All three of
  `_pop` / `_pop_polis` / `_pop_wp` are always kept, so this only changes the
  default column — you can pick a different denominator downstream without
  re-running.

Because `cfg` is built outside any `if (interactive())` guard (and behind a
`requireNamespace()` check), it is available both in RStudio and under
`Rscript`, so the starter scripts run headless.

## The starter scripts

```{r}
#| label: scripts
list.files(file.path(root, "02_scripts"))
```

`2a_download_data.R` pulls **every** input: the POLIS population reference, the
seven POLIS surveillance tables (`get_polis_data()`), and the WorldPop rasters
(`sntutils::download_worldpop*`). Only missing files are (re)fetched. The raw WHO
polio GDB shapefile layers are the one thing it does **not** download — drop
those into `01_data/1a_shapefiles/raw/` yourself.

`2b_process_data.R` runs in three sections, in order:

1. **Spatial** — `process_spatial()` cleans the GDB layers to
   `spatial_global_adm2.qs2` (everything else keys off it).
2. **WorldPop** — sums the annual rasters to adm2 tables (cached on disk;
   re-extracted only if missing), which `clean_pop()` then consumes.
3. **POLIS** — `run_pipeline(cfg = cfg)` cleans every stream (AFP / ES / SIA /
   LQAS / IM / virus / indicators **and** population via `clean_pop()`) and
   writes `polished_*` outputs plus `checks_*` workbooks.

## Running it end to end

```{r}
#| label: run
#| eval: false
# 0. drop the WHO polio GDB layers into 01_data/1a_shapefiles/raw/
# 1. download every input
source("02_scripts/2a_download_data.R")
# 2. process: shapefile -> worldpop extract -> run_pipeline over all streams
source("02_scripts/2b_process_data.R")
# 3. read a country / period slice of the cleaned outputs back
afg <- load_polished(country = "AFG")
names(afg)
```

`load_polished()` defaults `output_dir` to the active config, so once the
`.Rprofile` has run no path argument is needed — just the filter.

## Reproducible package versions with `renv`

Pass `renv = TRUE` to wire the project for
[renv](https://rstudio.github.io/renv/), so everyone installs the **same**
package versions:

```{r}
#| label: renv
#| eval: false
init_polis_pipeline("my_project", regions = "EMRO", renv = TRUE)
```

That runs `renv::scaffold()` — writing `renv/activate.R`, a starter `renv.lock`,
and a single clean autoloader line in the generated `.Rprofile`
(`source("renv/activate.R")`, no duplicate). The reproducibility loop is then:

1. install the packages the scripts need (`polished`, `sntutils`, `sf`, `qs2`,
   `terra`, `exactextractr`, …);
2. `renv::snapshot()` — pins those exact versions into `renv.lock` (renv finds
   them from the `pkg::fun` calls in `2a` / `2b` and the `.Rprofile`);
3. commit `renv.lock`;
4. a collaborator runs `renv::restore()` to reproduce the library exactly.

With `renv = FALSE` (the default) the `.Rprofile` still carries a *guarded*
autoloader — `if (file.exists("renv/activate.R")) source("renv/activate.R")` — so
the project is renv-ready: run `renv::init()` yourself later and nothing else in
the scaffold changes.

## Conventions baked in

- **Idempotent.** Re-running never clobbers an edited `.Rprofile` or script;
  existing files are kept unless you pass `overwrite = TRUE`. Directory creation
  is always safe to repeat.
- **`.here` sentinel.** A `.here` marker is written so `here::here()` in the
  `.Rprofile` resolves to the project root even before it is a git repo or
  RStudio project.
- **Sensitive raw data stays out of git.** The generated `.gitignore` excludes
  `01_data/**/raw/` — raw POLIS tables are individual-level line-lists (and
  large), so they should not be committed. Re-download them with `2a` instead.
- **One source of truth for paths.** The `.Rprofile` defines the roots; the
  scripts and `cfg` reference them, so there is never a second, drifting copy.

For a lighter, single-stream workspace (no `.Rprofile`, no scripts) see
`init_polis_project()` and the *End-to-end pipeline* article.
