---
title: "Computing surveillance indicators"
vignette: >
  %\VignetteIndexEntry{Computing surveillance indicators}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    df-print: default
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    message: false
    warning: false
---

```{r}
#| label: setup
#| include: false
library(polished)
```

Once the surveillance streams are cleaned (see the *Cleaning POLIS data* article),
`calc_polio_indicators()` computes the WHO POLIS indicator catalogue — the NPAFP
rate, stool adequacy, timeliness, dose history, environmental, virus, SIA and
composite families — at the admin levels you ask for.

## Browse the catalogue: `available_indicators()`

Every indicator the package knows about lives in a single registry. Read it with
`available_indicators()`, optionally filtered by `family`:

```{r}
#| label: catalogue
fams <- available_indicators()
table(fams$family)

# the AFP family, with what each indicator needs
available_indicators(family = "AFP")[, c("code", "label", "requires_pop")]
```

`requires_pop = TRUE` marks the indicators that need an under-15 population
denominator (the rate families); everything else is computed from the case/ES/
virus/SIA tables alone.

## Inputs

`calc_polio_indicators()` needs only a cleaned `cases` table to start; `es`,
`sia`, `virus`, `lab`, `population` and `admin_units` are optional and unlock
the indicators that depend on them. A missing source simply skips its
indicators with a message rather than erroring.

The example below builds a small **cleaned** AFP table by hand so it renders
without a live pull — in practice `cases` is the output of `clean_afp()` or
`run_pipeline()`.

```{r}
#| label: cases
cases <- tibble::tibble(
  epid = sprintf("NIE-BOR-D%02d-24-%03d", rep(1:2, length.out = 24), 1:24),
  classification_all = rep(c("NPAFP", "NPAFP", "NPAFP", "WPV1", "NOT-AFP", "PENDING"), 4),
  age_months = rep(c(12, 30, 48, 24), 6),
  year_onset = 2024L,
  paralysis_onset_date = as.Date("2024-01-01") + (1:24),
  adm0 = "NIGERIA", adm0_guid = "G0",
  adm1 = rep(c("BORNO", "YOBE"), length.out = 24),
  adm1_guid = rep(c("P1", "P2"), length.out = 24),
  adm2 = rep(c("MMC", "JERE", "GUJBA", "NGANZAI"), length.out = 24),
  adm2_guid = rep(c("D1", "D2", "D3", "D4"), length.out = 24),
  adequate_stool = rep(c("Yes", "No"), length.out = 24),
  notify_to_invest = rep(c(1, 2, 5, 1), 6)
)
```

The under-15 population denominator is an external table keyed by `adm2_guid`,
`year` and `u15_pop` (one row per admin unit per year):

```{r}
#| label: pop
population <- tibble::tibble(
  adm2_guid = c("G0", "P1", "P2", "D1", "D2", "D3", "D4"),
  year = 2024L,
  u15_pop = c(2e6, 1e6, 1e6, 5e5, 5e5, 5e5, 5e5)
)
```

## Compute

Ask for the admin levels you want; each is aggregated independently from the
case GUIDs. `indicators` selects what to compute: `"core"` (the default — the
core KPI set), `"all"` (the full registered catalogue), or an explicit vector of
codes. This article walks the broader catalogue, so it asks for `"all"`:

```{r}
#| label: calc
ind <- calc_polio_indicators(
  cases = cases,
  population = population,
  levels = c("adm0", "adm1", "adm2"),
  indicators = "all",
  verbose = FALSE
)
names(ind)
```

The result is a list: one **wide** tibble per level (`adm0` / `adm1` / `adm2`,
one row per admin unit × year with every indicator as a column), a tidy
**`long`** table (one row per admin × year × indicator), and a `meta` record of
what ran, was skipped, and the thresholds used.

```{r}
#| label: long
ind$long[
  ind$long$indicator %in% c("afp_count", "npafp_count", "npafp_rate", "stool_adequacy_pct"),
  c("level", "name", "indicator", "value")
]
```

The wide per-level tibble is the analyst-facing shape — one row per unit:

```{r}
#| label: wide
ind$adm1[, c("name", "afp_count", "npafp_count", "npafp_rate", "stool_adequacy_pct")]
```

## Population denominators and suppression

Rate indicators (e.g. `npafp_rate`) divide a count by the under-15 population
and scale by `rate_multiplier` (default per 100,000). To keep small-denominator
noise out of the output, a unit's rate is suppressed (`NA`) when its population
is below `min_pop` or its case count is below `min_cases`. Pass no `population`
and the rate families are skipped entirely — every population-independent
indicator (counts, percentages, timeliness) still computes:

```{r}
#| label: no-pop
no_pop <- calc_polio_indicators(
  cases = cases,
  levels = "adm0",
  indicators = "all",
  verbose = FALSE
)
no_pop$long[no_pop$long$indicator %in% c("npafp_count", "npafp_rate"), c("indicator", "value")]
```

### Choosing the population source

When the end-to-end pipeline builds the denominator itself from a raw POLIS
population table (via `clean_pop()`), you choose **which** population it uses
with `pop_source` on `polis_config()` (or directly on `clean_pop(pop_source =)`):

- `"reconciled"` (default): a trusted POLIS value, else WorldPop, else a
  district / province / country fallback.
- `"polis"`: the full POLIS population (WorldPop is ignored even if supplied).
- `"worldpop"`: the WorldPop value, else POLIS, else the fallback.

Whatever the mode, the cleaned population keeps `<age>_pop_polis` and
`<age>_pop_wp` next to the chosen `<age>_pop`, so the three sources stay
comparable.

See `?calc_polio_indicators` for the full argument set — the column-name
overrides, the classification vocabularies, and every threshold.
